---
title: "Example_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bbplot)
```

```{r}
library(here)
library(ggplot2)
library(tidytuesdayR)
library(dplyr)
library(lubridate)
library(forcats)
library(purrr)
```

You must describe what is the question you aim to answer with the data and data analysis.
This analysis aim to explore the Tornados in United States. 

You must describe and link to where the original data come from that you chose.
TThe dataset is from tidytuesday(https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-16/readme.md#tornados)

The data dictionary is [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-16/readme.md#data-dictionary)

```{r}
## Load Data
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

files = c("tornados.rda")
if (any(!file.exists(here("data", files)))) {
    tuesdata = tidytuesdayR::tt_load('2023-05-16')
    tornados = tuesdata$tornados
    # save the files to rda objects
    save(tornados, file = here("data", "tornados.rda"), compress = "xz")
}

load(here("data", "tornados.rda"))

```

```{r}
## glimpse of data 
map_chr(.x = tornados, .f = typeof)
## test NA
map(tornados, ~ any(is.na(.)))
```

```{r,fig.width=7}
FL=tornados%>%
  filter(st=="FL")%>%
  group_by(yr)%>%
  select(yr, mag)%>%
  summarise(fl_mag_mean=mean(mag, na.rm = T))
FL%>%
  ggplot(aes(x=yr, y=fl_mag_mean))+
  geom_line()+
  labs(title = "Florida average tornados magnitude from 1950-2020",
       subtitle = "There is a decreasing trend from 1950 to 2020. The lowest period is around 2000.",  x="years", y="average tornados magnitude", caption = "Chuyi Sun")+
  bbc_style()+
  theme(plot.title=element_text(size = 15))

```

```{r, fig.height=12, fig.width=10}
state_summary=tornados%>%
  mutate(tz=as.factor(tz))%>%
  mutate(st=as.factor(st))%>%
  select(-c("slat", "slon", "elat", "elon"))%>%
  group_by(st,mo)%>%
  arrange(mag)%>%
  summarise(state_mean=mean(mag, na.rm = T),.groups = "drop")
plot1=state_summary%>%
  ggplot(aes(x=st, y=state_mean))+
  geom_point()+
  facet_wrap(~mo, nrow = 6, scales = "free")+ 
  labs(title = "Average tornados magnitude of US states in each months",
       subtitle = "Except for September, other month trend to have similar average tornados magnitude. Alaska almost never have tornado", x="States", y="average tornados magnitude", caption = "Chuyi Sun")+
  geom_text(aes(label=st),size=2,vjust = -0.5)+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1, size = 5))

finalise_plot(plot_name = plot1,source = "Source: Example_analysis",save_filepath = "state_summary.png")
```


```{r}
## 1950-2022
## 1950-1974, 1975-1999, 2000-2024
tornados%>%
  mutate(stage=case_when(yr>1949 & yr<1975 ~ "1950-1974", yr>1974 & yr<2000 ~ "1975-1999",yr>1999 & yr<2025 ~ "2000-2024"))%>%
  select(stage, inj)%>%
  group_by(stage)%>%
  summarise(stage_sum=sum(inj))%>%
  ggplot(aes(x=stage, y=stage_sum))+
  geom_bar(stat="identity")+
  labs(title = "Total injury in tornados in every 24 years",
       subtitle = "Between 1950-1974, there are lot of injuries. And 2000-2024 has the lowest injuries", x="States", y="average tornados magnitude", caption = "Chuyi Sun")

```


## Functions used 

##### Dplyr  

* mutate
* group_by
* filter
* select
* case_when
* summarise


##### ggplot2  

* geom_line
* geom_point
* geom_bar
* geom_text
* facet_wrap

##### purrr 

* map_chr
* map 

##### bbplot  
* bbc_style
* finalise_plot




